---
title: ''
output:
  html_fragment:
    self_contained: false
    toc: false
    df_print: default
    mathjax: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<style>
h5 {
  font-size: 14pt;
  color: black;
  font_weight: normal;
}
h2 {
  font-size: 16pt;
  color: darkred;
  font_weight: normal;
}
h3 {
  font-size: 16pt;
  color: darkred;
  font_weight: normal;
}
p {
  font-size: 14pt;
  color: black;
  font_weight: normal;
}
</style>

## 🧠 Introduction and Motivation

Imagine a multicentric clinical trial designed to evaluate **three treatment strategies (A, B, and C)** for improving patient recovery scores.
Each participating **hospital (center)** treats patients with all three therapies.  

At first glance, it might seem that randomizing all patients across treatments—without considering centers—should suffice (a *Completely Randomized Design*, CRD).However, hospitals differ in many ways:  

- patient populations,  
- staff experience,  
- equipment,  
- and even environmental or procedural factors.  

These differences introduce **systematic variability** between centers that can **mask true treatment effects** if ignored.  
Treatments might appear less effective or more variable simply because centers differ — not because the therapies themselves do.

To handle this, we can treat **centers as blocks** and analyze the data using a **Randomized Complete Block Design (RCBD)**.  
In RCBD, each center applies all treatments, allowing us to separate:

- the variation **due to centers** from  
- the variation **due to treatments**,  

thereby improving the **precision** and **power** of statistical conclusions.

In this exercise, you will use the **RCBD simulation Shiny app** to explore how experimental design choices (such as block effect, within-center variability, number of centers, and replicates) influence the conclusions drawn from RCBD and CRD analyses.

> **Key questions to explore**

> 1. How does accounting for center effects improve inference in multicentric studies?  
> 2. Under what conditions does blocking yield substantial gains in precision or power?  
> 3. When might RCBD and CRD lead to similar results?

---

## 🧩 Learning Objectives

By completing this exercise, you should be able to:

- Understand the conceptual role of **blocking** in multicenter studies.  
- Evaluate how **design parameters** (variability, number of blocks, replicates, and effect size) affect statistical **power** and **precision**.  
- Compare **ANOVA outputs** for RCBD vs CRD and interpret the implications.  
- Identify efficient experimental setups that balance power and resource use.

---

## Part 1 — Baseline Scenario

**Set:**

- Baseline = **100**
- `sigma_within` = **5**
- Cohen's *f* = **0.25** (medium)
- Centers (blocks) = **6**
- Replicates per treatment per center = **5**
- Block effect = **Medium**

### 1.1 Record results

```{r, echo=FALSE}
baseline_tbl <- data.frame(
  Model = c("RCBD", "CRD"),
  F_treatment = "",
  p_treatment = "",
  F_block = c("", ""),
  p_block = c("", ""),
  MSE = "",
  Power = "",
  Avg_CI_Width_for_Pairwise_Diffs = ""
)
knitr::kable(baseline_tbl, caption = "Baseline results (to be filled in).")
```

### 1.2 Interpretation

- Does including centers as blocks reduce residual variance (MSE)?
- How does it affect treatment significance and **power**?
- Why could CRD be misleading in a multicentric setting?

---

## Part 2 — Effect of Between-Center Heterogeneity (Block Effect)

Keep parameters as in Part 1, **vary only the Block effect**: *None*, *Low*, *Medium*, *High*.

```{r, echo=FALSE}
block_levels <- c("None","Low","Medium","High")
results_blocks <- data.frame(
  Block_Effect = rep(block_levels, each = 2),
  Model = rep(c("RCBD","CRD"), times = length(block_levels)),
  F_treatment = "",
  p_treatment = "",
  MSE = "",
  Power = "",
  Avg_CI_Width_for_Pairwise_Diffs = ""
)
knitr::kable(results_blocks, caption = "Results across block-effect levels (fill in from app).")
```

### Interpretation

- When block effects are absent, does blocking help or hurt power? Why?
- When block effects are strong, how do **MSE**, **power**, and **treatment p-values** compare between RCBD and CRD?

---

## Part 3 — Effect of Within-Center Variability (sigma_within)

Fix parameters as in Part 1, set **Block effect = Medium**. Vary `sigma_within`:
**2, 5, 10, 20**.

```{r, echo=FALSE}
sigma_levels <- c(2,5,10,20)
results_sigma <- data.frame(
  sigma_within = rep(sigma_levels, each = 2),
  Model = rep(c("RCBD","CRD"), times = length(sigma_levels)),
  F_treatment = "",
  p_treatment = "",
  MSE = "",
  Power = "",
  Avg_CI_Width_for_Pairwise_Diffs = ""
)
knitr::kable(results_sigma, caption = "Effect of within-center variability (fill in).")
```

### Interpretation
- How do **power** and **CI widths** change as `sigma_within` increases?
- Can blocking compensate when residual noise is very large? Explain.

---

## Part 4 — Number of Centers and Replicates

Fix: Baseline = 100, `sigma_within` = 5, Cohen's *f* = 0.25, Block effect = *Medium*.
Explore combinations:

- Centers = **3, 6, 9**
- Replicates = **2, 4, 6**

```{r, echo=FALSE}
grid <- expand.grid(
  Centers = c(3,6,9),
  Replicates = c(2,4,6),
  Model = c("RCBD","CRD"),
  KEEP.OUT.ATTRS = FALSE,
  stringsAsFactors = FALSE
)
grid$Power <- ""
grid$Avg_CI_Width_for_Pairwise_Diffs <- ""
knitr::kable(grid, caption = "Power & precision over centers × replicates (fill in).")
```

### Interpretation
- Identify **diminishing returns** when increasing replicates or centers.
- Propose an **efficient** design achieving **power ≥ 0.80** with minimal sample size.

---

## Part 5 — Precision of Treatment Differences

Choose two contrasting scenarios (e.g., *Low* vs *High* block effect). For **RCBD**:
- Report the **pairwise estimates (A–B, A–C, B–C)** and **95% CIs**.
- Summarize how blocking impacts **CI width**.

```{r, echo=FALSE}
pairs_tbl <- data.frame(
  Contrast = c("A - B","A - C","B - C"),
  Estimate = "",
  CI_lower = "",
  CI_upper = "",
  Scenario = ""
)
knitr::kable(pairs_tbl, caption = "Pairwise treatment differences (RCBD).")
```

### Short discussion
- Under which conditions do CIs shrink the most with RCBD?
- Are there situations where CRD and RCBD yield similar precision? Why?

---

## Part 6 — Reflection (Short Answers)

1. When do **centers as blocks** provide the largest advantage?  
2. How do `sigma_within`, number of centers, and replicates jointly determine **power**?  
3. What’s a **practical combination** of centers × replicates to achieve power ≥ 0.80 in this context?  
4. If the block effect were negligible, what would be the consequence of using RCBD instead of CRD?  
5. How would you check the key assumption that **treatment effects are consistent across centers** (i.e., no treatment×center interaction)?

---

## Optional: Notes for Instructors (remove before submission)

- Expected trends:  
  - **Higher block effect** → RCBD gains over CRD in power and MSE.  
  - **Higher sigma_within** → lower precision and power for both; RCBD helps mainly when block variance is non-negligible.  
  - **More centers** controls systematic heterogeneity; **more replicates** reduces within-center error.  
  - **Diminishing returns** after moderate replication.

- Suggested grading criteria: completeness of tables; correctness of interpretations; ability to justify an efficient design.
